---
title: "Topology-aware clustering of longitudinal trajectories with LongOmicsTraj"
author: "Nazanin Zounemat-Kermani"
package: LongOmicsTraj
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Topology-aware clustering of longitudinal trajectories}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
```

# Overview

LongOmicsTraj separates longitudinal trajectory estimation from trajectory
representation and clustering. In this vignette, the packaged GLUCOLD example
is used to demonstrate a complete topology-aware clustering workflow:

1. construct visit-level empirical trajectories;
2. calculate adjacent changes and assign feature-level Omics Trajectory
   Signatures (OTS);
3. cluster genes using combined group-by-visit profiles;
4. summarise topology composition and purity within clusters;
5. evaluate topology coverage and duplication;
6. visualise cluster centres, trajectory arms and topology deviations; and
7. fit topology-aware FlexMix trajectory clusters in a selected treatment arm.

The example is intentionally small so that it can be run as part of package
documentation. The packaged object contains 71 molecular features measured
across three visits and four treatment groups.

# Required packages

```{r packages}
library(LongOmicsTraj)
library(S4Vectors)
library(SummarizedExperiment)
library(MultiAssayExperiment)
library(dplyr)
library(ggplot2)
```

The optional sections use `cluster`, `matrixStats`, `patchwork`, `scales`,
`withr` and `flexmix`.

# Load and prepare the GLUCOLD example

```{r load-data}
data("lot_glucold")

visits <- c(
  "Baseline",
  "6 Months",
  "30 Months"
)

assay_name <- "transcriptomics"

lot <- lot_glucold

experiments <- MultiAssayExperiment::experiments(
  lot@mae
)

se <- experiments[[assay_name]]

metadata <- as.data.frame(
  SummarizedExperiment::colData(se)
)

metadata$visit <- factor(
  metadata$visit,
  levels = visits,
  ordered = TRUE
)

metadata$group <- factor(
  metadata$group
)

stopifnot(
  !anyNA(metadata$visit)
)

SummarizedExperiment::colData(se) <- S4Vectors::DataFrame(
  metadata,
  row.names = rownames(metadata)
)

experiments[[assay_name]] <- se

lot@mae <- MultiAssayExperiment::MultiAssayExperiment(
  experiments = experiments
)

methods::validObject(lot)

lot
```

The visit order is set explicitly because topology labels depend on the
chronological order of adjacent transitions.

# Construct feature-level trajectories

For this vignette, empirical group-by-visit means are used. The same clustering
workflow can be applied after constructing a VMM with `lot_from_lmm()`,
`lot_from_gam()` or `lot_from_masigpro()`.

```{r empirical-vmm}
lot_features <- lot_from_empirical(
  object = lot,
  assay = assay_name,
  visits = visits,
  group_col = "group",
  visit_col = "visit",
  estimator = "empirical",
  overwrite = TRUE,
  verbose = FALSE
)

utils::head(
  as.data.frame(lot_features@vmm)
)
```

# Map feature trajectories to OTS labels

Adjacent visit-to-visit changes are calculated once. The example uses the
hybrid thresholding strategy only as a reproducible demonstration; the
threshold should be selected and examined through sensitivity analysis in
substantive applications.

```{r feature-ots}
lot_features <- lot_compute_deltas(
  object = lot_features,
  assay = assay_name,
  visits = visits,
  overwrite = TRUE
)

lot_features <- lot_compute_thresholds(
  object = lot_features,
  assay = assay_name,
  method = "hybrid",
  visits = visits,
  overwrite = TRUE,
  verbose = FALSE
)

lot_features <- lot_map_to_ots(
  object = lot_features,
  assay = assay_name,
  visits = visits,
  overwrite = TRUE,
  verbose = FALSE
)

table(
  lot_features@ots$topology_label
)
```

Each OTS is an ordered pair of transition states because GLUCOLD contains
three visits. For example, `up_down` denotes an increase from baseline to
6 months followed by a decrease from 6 to 30 months.

# Cluster genes across all treatment groups

`lot_from_clusters()` represents each gene by its complete set of
group-by-visit means. Clustering therefore reflects combined longitudinal
behaviour across all treatment arms rather than a single endpoint or group.

The example uses k-means with four clusters and standardises each gene
trajectory before clustering.

```{r kmeans-clustering}
lot_clusters <- lot_from_clusters(
  object = lot_features,
  assay = assay_name,
  group_col = "group",
  visit_col = "visit",
  visits = visits,
  k = 4L,
  estimator = "kmeans",
  scale_features = TRUE,
  seed = 1L,
  require_gene_ots = TRUE,
  overwrite = TRUE,
  verbose = FALSE
)

lot_clusters
```

The cluster memberships, cluster sizes, cluster profiles and clustering
diagnostics are stored in the `results` slot.

```{r inspect-clustering-results}
as.data.frame(
  lot_clusters@results$cluster_sizes
)

as.data.frame(
  lot_clusters@results$cluster_diagnostics
)

utils::head(
  as.data.frame(
    lot_clusters@results$cluster_membership
  )
)
```

The VMM now contains cluster-centre trajectories rather than gene-level
trajectories.

```{r inspect-cluster-vmm}
utils::head(
  as.data.frame(lot_clusters@vmm)
)
```

# Map cluster centres to topology labels

Cluster centres can be processed by the same delta, threshold and OTS
functions used for molecular features.

```{r cluster-centre-ots}
lot_clusters <- lot_compute_deltas(
  object = lot_clusters,
  assay = assay_name,
  visits = visits,
  overwrite = TRUE
)

lot_clusters <- lot_compute_thresholds(
  object = lot_clusters,
  assay = assay_name,
  method = "hybrid",
  visits = visits,
  overwrite = TRUE,
  verbose = FALSE
)

lot_clusters <- lot_map_to_ots(
  object = lot_clusters,
  assay = assay_name,
  visits = visits,
  overwrite = TRUE,
  verbose = FALSE
)

as.data.frame(
  lot_clusters@ots
)
```

These cluster-centre OTS labels describe the representative trajectory of each
cluster. They should be interpreted together with the feature-level topology
composition stored during clustering.

# Inspect topology composition within clusters

When feature-level OTS labels are present before clustering,
`lot_from_clusters()` stores the within-cluster topology distribution in
`cluster_ots_distribution`.

```{r cluster-ots-distribution}
cluster_ots_distribution <- as.data.frame(
  lot_clusters@results$cluster_ots_distribution
)

utils::head(
  cluster_ots_distribution
)
```

## Trajectory-arm plot

The trajectory-arm plot shows that a conventional molecular-feature cluster
may contain several distinct OTS classes. Arrow width is proportional to the
within-cluster topology proportion.

```{r trajectory-arms, fig.cap="Feature-level trajectory topologies within each cluster and treatment group. Arrow width represents the within-cluster topology proportion."}
lot_plot_cluster_trajectory_arms(
  object = lot_clusters,
  top_n = 5L,
  min_proportion = 0.05,
  show_counts = TRUE,
  show_percent = TRUE,
  scale_width = TRUE,
  ncol = 2
)
```

# Quantify topology purity

Topology purity is the proportion of genes assigned to the dominant OTS within
each cluster-group combination. Entropy provides a complementary measure of
within-cluster topology heterogeneity.

```{r cluster-purity}
purity_assignments <- as.data.frame(
  lot_clusters@results$gene_ots
) |>
  dplyr::transmute(
    gene_id = as.character(object_id),
    cluster = as.character(cluster_id),
    group = as.character(group),
    topology_label = as.character(topology_label)
  )

purity <- lot_cluster_purity(
  assignments = purity_assignments,
  purity_threshold = 0.90
)

purity$cluster_group_purity

purity$overall
```

A purity of one means that all genes in a cluster-group combination have the
same OTS. Lower purity indicates that the cluster combines several trajectory
topologies.

# Calculate topology coverage and duplication

High purity alone does not guarantee that a clustering solution represents
distinct longitudinal behaviours. Several pure clusters may still have the
same dominant topology. `lot_cluster_topology_score()` therefore reports
topology coverage and duplicate topology counts separately from purity.

```{r topology-score}
cluster_diagnostics <- as.data.frame(
  lot_clusters@results$cluster_diagnostics
)

minimum_cluster_size <- min(
  as.data.frame(
    lot_clusters@results$cluster_sizes
  )$n_features
)

topology_score <- lot_cluster_topology_score(
  purity = purity,
  k = cluster_diagnostics$k[[1L]],
  minimum_cluster_size = minimum_cluster_size
)

topology_score$group_summary

topology_score$topology_repetition

topology_score$overall
```

Topology coverage is the number of unique dominant topologies divided by the
number of observed clusters within each group. Values below one indicate that
multiple clusters share the same dominant temporal behaviour.

# Plot cluster-centre trajectories

Cluster profiles are available in both raw and scaled form.

```{r cluster-centre-plot, fig.cap="Scaled cluster-centre trajectories across treatment groups and visits."}
cluster_profiles <- as.data.frame(
  lot_clusters@results$cluster_profiles
) |>
  dplyr::filter(
    value_type == "scaled"
  )

ggplot(
  cluster_profiles,
  aes(
    x = visit,
    y = value,
    group = group,
    colour = group
  )
) +
  geom_line(
    linewidth = 0.9
  ) +
  geom_point(
    size = 1.8
  ) +
  facet_wrap(
    ~ cluster_id,
    scales = "free_y"
  ) +
  labs(
    title = "Cluster-centre longitudinal profiles",
    x = "Visit",
    y = "Scaled cluster centre",
    colour = "Group"
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(
      angle = 30,
      hjust = 1
    ),
    legend.position = "bottom"
  )
```

# Alternative clustering with PAM

`lot_from_clusters()` also supports partitioning around medoids. PAM is useful
when a medoid-based solution is preferred and the optional `cluster` package is
available.

```{r pam-clustering, eval=requireNamespace("cluster", quietly=TRUE)}
lot_pam <- lot_from_clusters(
  object = lot_features,
  assay = assay_name,
  group_col = "group",
  visit_col = "visit",
  visits = visits,
  k = 4L,
  estimator = "pam",
  scale_features = TRUE,
  require_gene_ots = TRUE,
  overwrite = TRUE,
  verbose = FALSE
)

as.data.frame(
  lot_pam@results$cluster_diagnostics
)
```

# Topology-aware FlexMix clustering within one treatment arm

`lot_from_flexmix_clusters()` fits finite mixtures of longitudinal regression
trajectories within a selected group. Candidate values of `k` are compared
using BIC, topology purity, the fraction of clusters passing the purity rule
and minimum cluster size.

The withdrawal arm is used here because it is expected to contain biologically
meaningful non-monotonic responses.

```{r flexmix-clustering, eval=requireNamespace("flexmix", quietly=TRUE)}
withdrawal_group <- "ICS 6 months then withdrawal"

lot_flexmix <- lot_from_flexmix_clusters(
  object = lot_features,
  assay = assay_name,
  group = withdrawal_group,
  visits = visits,
  k = 2:5,
  group_col = "group",
  visit_col = "visit",
  sample_col = "sample_id",
  trajectory_source = "estimated",
  trajectory_estimator = "empirical",
  model = "quadratic",
  scale_features = TRUE,
  purity_threshold = 0.90,
  required_passing_fraction = 0.80,
  min_cluster_size = 5L,
  nrep = 5L,
  seed = 42L,
  store_fits = FALSE,
  overwrite = TRUE,
  verbose = FALSE
)

as.data.frame(
  lot_flexmix@results$flexmix_model_comparison
)

lot_flexmix@results$flexmix_selected_k

lot_flexmix@results$flexmix_selection_reason
```

The selected resolution is the smallest candidate satisfying the topology
purity and cluster-size criteria. If none satisfies both criteria, the
lowest-BIC solution is selected.

## Plot the cluster-selection path

```{r cluster-selection-plot, eval=requireNamespace("flexmix", quietly=TRUE), fig.cap="Likelihood-based and topology-aware statistics across candidate FlexMix cluster resolutions."}
lot_plot_cluster_selection(
  object = lot_flexmix,
  metrics = c(
    "BIC",
    "weighted_purity",
    "passing_fraction",
    "minimum_cluster_size"
  ),
  purity_threshold = 0.90,
  required_passing_fraction = 0.80,
  min_cluster_size = 5L,
  show_selected = TRUE
)
```

## Plot representative cluster trajectories and OTS composition

```{r cluster-topology-plot, eval=requireNamespace("flexmix", quietly=TRUE) && requireNamespace("patchwork", quietly=TRUE), fig.cap="Representative FlexMix cluster trajectories and their feature-level OTS composition."}
lot_plot_cluster_topology(
  object = lot_flexmix,
  group = withdrawal_group,
  value_type = "scaled",
  show_features = TRUE,
  max_feature_lines = 30L,
  purity_threshold = 0.90,
  sort_by = "purity",
  seed = 1L
)
```

## Plot topology deviations

The topology-deviation plot compares the frequency of each OTS within a cluster
with its overall frequency among all clustered genes. Positive deviations
indicate enrichment and negative deviations indicate depletion.

```{r topology-deviation-plot, eval=requireNamespace("flexmix", quietly=TRUE), fig.cap="Enrichment or depletion of OTS classes within selected FlexMix clusters relative to the full clustered feature set."}
lot_plot_topology_deviation(
  object = lot_flexmix,
  group = withdrawal_group,
  min_abs_deviation = 0.02,
  top_n = 6L
)
```

# Plot representative genes

Genes assigned to selected FlexMix clusters can be retrieved from the stored
assignment table and visualised using the original longitudinal measurements.

```{r representative-genes, eval=requireNamespace("flexmix", quietly=TRUE)}
assignments <- as.data.frame(
  lot_flexmix@results$flexmix_gene_assignments
)

representative_genes <- assignments |>
  dplyr::filter(
    !is.na(topology_label)
  ) |>
  dplyr::group_by(
    cluster
  ) |>
  dplyr::slice_head(
    n = 1L
  ) |>
  dplyr::ungroup() |>
  dplyr::pull(
    gene_id
  )

representative_genes
```

```{r representative-gene-plot, eval=requireNamespace("flexmix", quietly=TRUE), fig.cap="Observed longitudinal measurements for one representative gene from each selected FlexMix cluster."}
lot_plot_trajectory(
  object = lot_features,
  features = representative_genes,
  assay = assay_name,
  groups = withdrawal_group,
  estimator = "empirical",
  show_subjects = TRUE,
  summary = "mean_se",
  facet_scales = "free_y"
)
```

# Interpretation

This workflow separates three related but distinct questions:

1. **Do genes have similar numerical profiles?**  
   Addressed by k-means, PAM or FlexMix clustering.

2. **Do members of a cluster share the same temporal topology?**  
   Addressed by topology purity and entropy.

3. **Do different clusters represent distinct temporal behaviours?**  
   Addressed by topology coverage and duplicate topology counts.

A clustering solution may have favourable likelihood or compactness while
still merging several OTS classes, or it may split one dominant topology into
several clusters. LongOmicsTraj therefore reports conventional clustering
diagnostics and topology-aware summaries separately rather than combining them
into a single unvalidated score.

# Session information

```{r session-info}
sessionInfo()
```
